---
title: "R_Project_DSE5002"
author: "Isabella Parlato"
date: "2024-11-18"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

#Objective

Your CEO has decided that the company needs a full-time data scientist, and possibly a team of them in the future. She thinks she needs someone who can help drive data science within then entire organization and could potentially lead a team in the future. She understands that data scientist salaries vary widely across the world and is unsure what to pay them. To complicate matters, salaries are going up due to the great recession and the market is highly competitive. Your CEO has asked you to prepare an analysis on data science salaries and provide them with a range to be competitive and get top talent. The position can work offshore, but the CEO would like to know what the difference is for a person working in the United States. Your company is currently a small company but is expanding rapidly. 

Prepare your analysis in an R file. Your final product should be a power point presentation giving your recommendation to the CEO. CEOs do not care about your code and don’t want to see it. They want to see visuals and a well thought out analysis. You will need to turn in the power point and the code as a flat R file. 

#Restating the questions
What is the competitive global pay range for a Full-Time Data Scientist position? 
What is pay difference for fully offshore remote worker vs a US only employee (keeping in mind that a fully remote person can work offshore)?

#Add data into r

```{r}
library(readr)
raw_ds_salaries <- read_csv("data/r project data.csv")

head(raw_ds_salaries)

```

# initial cleaning

```{r}
#add packages

library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)


```


```{r}
#want to remove columns for "salary" and "salary_currency" & then need to narrow down to only DS position and FT based on brief

ft_ds_only_salaries <- raw_ds_salaries %>%
  select(-salary, -salary_currency) %>%
  filter(job_title == "Data Scientist") %>%
  filter(employment_type == "FT") %>%
  select(-employment_type, -job_title, -"...1")

head(ft_ds_only_salaries)

colnames(ft_ds_only_salaries) #used this to figure out name of first column in raw data since it was just numbers so went back up to remove it from cleaned up data

```

```{r}
# now i want to clean up how some of the character variables names

ft_ds_only_salaries <- ft_ds_only_salaries %>% 
  mutate(company_size = recode(company_size, "L" = "Large", "M" = "Medium", "S" = "Small"))

ft_ds_only_salaries <- ft_ds_only_salaries %>% mutate(experience_level = recode( experience_level, "EN" = "Entry-level", "MI" = "Junior Mid-level", "SE" = "Intermediate Senior-level", "EX" = "Expert Executive-level"))

 
```


```{r}
#trying to fix order in which positions are pulling
unique(ft_ds_only_salaries$experience_level)

ft_ds_only_salaries$experience_level <- factor(ft_ds_only_salaries$experience_level, level = c("Entry-level", "Junior Mid-level", "Intermediate Senior-level"))

```



```{r}
summary_ft_ds_only <- summary(ft_ds_only_salaries)
print(summary_ft_ds_only)

```

Based on summary, regarding salary_in_usd for only Data Science Roles, regardless of other factors:

1st Qu.: $55,490.00 this is the median of the lower half
3rd Qu.: $141,975.00 this is the median of the upper half
Median : $104,796.00
Mean   : $108,923.00
Min.   : $2,859.00  this feels like a pretty extreme outlier to have 
Max.   : $412,000.00 similar to the min., this is also a significant outlier

```{r}
#IQR of ft_ds_only_salaries based on summary data

iqr_ft_ds_only_salaries <- 141975.00 - 55490.00

# $86,485.00 interquartile range
```



```{r}
#next i want to try separating out by company size

ft_ds_small_companies <- ft_ds_only_salaries %>%
  filter(company_size == "Small")
head(ft_ds_small_companies)


ft_ds_medium_companies <- ft_ds_only_salaries %>%
  filter(company_size == "Medium")
head(ft_ds_medium_companies)


ft_ds_large_companies <- ft_ds_only_salaries %>%
  filter(company_size == "Large")
head(ft_ds_large_companies)

```

```{r}
#summaries & IQR by company size for ds_only_salaries

summary(ft_ds_small_companies)
iqr_ft_ds_small_companies_salaries <- 83810 - 23375
print(iqr_ft_ds_small_companies_salaries)
# $60,435.00 IQR for small companies

summary(ft_ds_medium_companies)
iqr_ft_ds_medium_companies_salaries <- 165110 - 88352
print(iqr_ft_ds_medium_companies_salaries)
# $76,758.00 IQR for medium companies

summary(ft_ds_large_companies)
iqr_ft_ds_large_companies_salaries <- 135000 - 50000
print(iqr_ft_ds_large_companies_salaries)
# $85,000.00 IQR for large companies

#need to keep in mind that this looks at salary for company size regardless of year and also regardless of experience

```


next I want to try to create a plot to show a visualization of salary range divided by company size and showing separate years and a breakdown by experience level 

```{r}
#trying to see how it looks with turkeyplot

ft_ds_only_salaries %>%
  ggplot(mapping = aes(x = experience_level, y = salary_in_usd, color = company_size)) + geom_boxplot(outlier.shape = NA, varwidth = T) +
  geom_jitter(position = position_jitterdodge (0.5), size = 0.75) +
  facet_grid(company_size~.) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x='Experience Level of Employee'
      ,y='Salary in USD'
      ,color = 'Company Size'
      ,title='Salaries For Full-Time Data Scientists: On A Global Scope')
  

```

```{r}
# want to see how it looks as a barchart as well with year factored in

ggplot(ft_ds_only_salaries) +
  geom_col(mapping= aes(x = experience_level, y = salary_in_usd, color = company_size)) +   facet_grid(company_size ~ work_year, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 30)) +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x='Experience Level of Employee'
      ,y='Salary in USD'
      ,title='Salaries For Full-Time Data Scientists: On A Global Scope')


```

with how the above chart came out, it seems like there isn't much data in some years vs others so we're likely better off not separating by year with the idea that salaries will increase over time 

```{r}
# trying without year factored in

ggplot(ft_ds_only_salaries) +
  geom_col(mapping= aes(x = experience_level, y = salary_in_usd, color = company_size)) +  facet_grid(company_size~., scales = "free_y") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  theme(axis.text.x = element_text(angle = 10)) +
  labs(x='Experience Level of Employee'
      ,y='Salary in USD'
      ,title='Salaries For Full-Time Data Scientists: On A Global Scope')

#after comparing all three plots, I think turkeyplot seems like least skewed view - use that in slide and need to accompany with table with min/max/IQRs for clarity

```


```{r}
#now i need to narrow the DS only positions table down from all global ft to us based workers and full time remote off shore workers, which would allow them to work at our us based company. I'll split it into two data sets: us only and then non-us fully remote then I'll need to join them into one table

us_only_ft_ds_salaries <- ft_ds_only_salaries %>%
  filter(employee_residence == "US")
head(us_only_ft_ds_salaries)

summary(us_only_ft_ds_salaries)
# median: $140,000 - mean: $149,408 - 3rd Q: $174,500 - 1st Q: $120,000 - iqr: $54,500

```

```{r}

non_us_fully_remote_salaries <- ft_ds_only_salaries %>%
  filter(employee_residence != "US") %>%
  filter(remote_ratio == "100")

head(non_us_fully_remote_salaries)

summary(non_us_fully_remote_salaries)
# median: $45,760 - mean: $51,046 - 3rd Q: $69,336 - 1st Q: $31,615 - iqr: $37,721
```


```{r}
#since the ceo wants to know difference between US workers and that of offshore

non_us_salaries_ft_ds <- ft_ds_only_salaries %>%
  filter(employee_residence != "US")

head(non_us_salaries_ft_ds)

summary(non_us_salaries_ft_ds)
# median: $49,823 - mean: $57989 - 3rd Q: $79,296 - 1st Q: $35,962 - iqr: $43,334

```



```{r}

full_us_offshore_ft_ds_salaries <- us_only_ft_ds_salaries %>%
  full_join(non_us_fully_remote_salaries)

head(full_us_offshore_ft_ds_salaries)

```
 
```{r}
summary(full_us_offshore_ft_ds_salaries)
```

salary stats from summary: min - 5,679.00; max - 412,000.00; median - 130,000.00; mean - 125,534.00; 1st Q - 81,250.00; 3rd Q - 160,000.00

```{r}
# IQR for us eligible ds salaries only

iqr_us_offshore_ft_ds_salaries <- 160000.00 - 81250.00
#$78,750

```


```{r}
# now to create a visualization for this 

ggplot(full_us_offshore_ft_ds_salaries, mapping= aes(x = experience_level, y = salary_in_usd, color = company_size)) +
 geom_bar (stat = 'identity', position = 'dodge')   +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  labs(x='Experience Level of Employee'
      ,y='Salary in USD'
      ,color = 'Company Size'
      ,title='Salaries For Full-Time Data Scientists: US-Based & Fully-Remote Offshore')

  
```

```{r}
#now to pull summaries by company size again to get IQR since it seems like there are some outliers involved

large_us_remote_offshore_summary <- full_us_offshore_ft_ds_salaries %>%
  filter (company_size == "Large")
summary(large_us_remote_offshore_summary)


iqr_large_us_remote_offshore_summary <- 149800 - 56599
#IQR large companies: $93,201.00


medium_us_remote_offshore_summary <- full_us_offshore_ft_ds_salaries %>%
  filter (company_size == "Medium")
summary(medium_us_remote_offshore_summary)

iqr_medium_us_remote_offshore_summary <- 171500 - 114723
#IQR medium companies: $56,777.00

    
small_us_remote_offshore_summary <- full_us_offshore_ft_ds_salaries %>%
  filter (company_size == "Small")
summary(small_us_remote_offshore_summary)

iqr_small_us_remote_offshore_summary <- 88869 - 35646
#IQR small companies: $53,223.00 

```

#looking ahead/expanding beyond for perspective

It feels like looking solely at DS only positions isn't much data so in an effort to help give the CEO perspective about salaries for data roles, I want to pull the 5 most popular job positions plus DS if it isn't one of them from the raw data, once again limit down to FT only since that's the type of role that she wants. I also think I can remove large companies since we're a small company on our way to becoming medium sized.

```{r}
#figure out top 5 positions

raw_ds_salaries %>% count(job_title)

# Data Analyst - 97; Data Engineer - 132; Data Scientist - 143; Machine Learning Engineer - 41; Research Scientist - 16

```

```{r}
#now to clean up df to just those 5 positions and remove large companies and clean up column variables names

top_five_positions_global_no_large_companies <- raw_ds_salaries %>%
  select(-salary, -salary_currency) %>%
  filter(job_title %in% c("Data Scientist", "Data Analyst", "Data Engineer", "Machine Learning Engineer", "Research Scientist")) %>%
  filter(employment_type == "FT") %>%
  select(-employment_type, -"...1") %>%
  filter (company_size != "L")

top_five_positions_global_no_large_companies <- top_five_positions_global_no_large_companies %>% 
  mutate(company_size = recode(company_size, "M" = "Medium", "S" = "Small"))

top_five_positions_global_no_large_companies <- top_five_positions_global_no_large_companies %>% mutate(experience_level = recode( experience_level, "EN" = "Entry-level", "MI" = "Junior Mid-level", "SE" = "Intermediate Senior-level", "EX" = "Expert Executive-level"))

head(top_five_positions_global_no_large_companies)

```

```{r}

summary(top_five_positions_global_no_large_companies)

#mean: 108702 - median: 102100 - 3rd Q: 140000 - 1st Q: 65949 - IQR: 74051

```

```{r}
#trying to fix order in which positions are pulling so correct on plot

unique(top_five_positions_global_no_large_companies$experience_level)

top_five_positions_global_no_large_companies$experience_level <- factor(top_five_positions_global_no_large_companies$experience_level, level = c("Entry-level", "Junior Mid-level", "Intermediate Senior-level", "Expert Executive-level"))

```



```{r}
#now to create a visualization

ggplot(top_five_positions_global_no_large_companies, mapping= aes(x = experience_level, y = salary_in_usd, color = job_title)) +
 geom_bar (stat = 'identity', position = 'dodge')   +
  facet_grid(company_size~., scales = "free_y") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",")) +
  theme(axis.text.x = element_text(angle = 10)) +
  labs(x='Experience Level of Employee'
      ,y='Salary in USD'
      ,color = "Job Position Title"
      ,title='Global Salaries For Full-Time Top-Five Data Science Positions')


```

```{r}
#now to break it down by company size as well

small_companies_top_five_global <- top_five_positions_global_no_large_companies %>%
  filter (company_size == "Small")
summary(small_companies_top_five_global)

iqr_small_companies_top_five_global <- 83125 - 42070
#median: $61,363 - mean: $61,416 - IQR: $41,055
  

medium_companies_top_five_global <- top_five_positions_global_no_large_companies %>%
  filter (company_size == "Medium")
summary(medium_companies_top_five_global)

iqr_medium_companies_top_five_global <- 150000 - 78526
#median: $113,950 - mean: $117,499 - IQR: $71,474

```







